Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular cross sections

### 2011-09-12 Text/Submission/paper.tex

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{Figures/capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{Figures/inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{Figures/legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

@@

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which would require a
simultaneous imposement of pressure boundary conditions and the body force. 
%The pressure boundary
%conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
%horizontal direction only and the upward/downward flows with pressure difference imposed are not
%simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

@@

To further examine the steady-state, the velocity in the streamwise direction is plotted. The values
of the velocity are taken on the contour where $\phi=0$. This corresponds
 to the bubble interface. That allows to check whether the bubble is propagating as a whole rigid
body or it's shrinking or elongating. This is one of the characteristics of the steady-state.  One
can see in Fig. \ref{fig:velocity:contour} the contour plot and the
corressponing streamwise velocity
component. The crossection is a plane $x=0$, where $x$ points
towards the reader, $z$ is the streamwise direction. As fluxes inside the bubble have
clockwise and counterclockwise directions, see Section \ref{sec:velocity}, one needs to thoroughly
examine only the points corresponding to the front and the rear bubble caps. If they are equal, then
it is exactly the velocity with which the bubble propagates in the microchannel as a rigid body.
For a physical analogy, one can imagine the rotating ball propagating in the streamwise direction.
The front and rear ball points have the same streamwise velocity. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{Figures/velocity_interface_contour.eps}\\
\includegraphics[width=\textwidth]{Figures/velocity_interface_values.eps}\\
\caption{The bubble shape and the corresponding streamwise velocities taken on the bubble
interface after $240000$ iterations. Directions $y$ and $z$ are scaled to the half of the channel
$H_{\mathrm{eff}}/2$. One can see that the bubble front and rear caps are propagating with nearly
the
same velocity, that is the actual bubble velocity. For more details on velocity patterns
see Section \ref{sec:velocity}. \label{fig:velocity:contour}}
\end{figure}

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

\subsection{Velocity pattern}
\label{sec:velocity}
The liquid velocity pattern is known to change its behavior depending on the capillary number. For
relatively low capillary numbers $Ca<0.6$, in a reference frame moving with the bubble a vortex 
is observed \cite{heil-threedim}. For larger capillary numbers it is indicated that there is no
vortex in front of the bubble. The transition capillary number specified by
\citet{heil-threedim} for a square channel as $Ca=0.691$. However, the
distribution of vorticity strongly depends on the Reynolds number \cite{heil-bretherton}, as well as
on
the
pressure distribution. In the case of the bubble train it is indicated by \citet{kreutzer-taylor}
that the pressure is significantly influenced by bubble frequency and slug distance. The present
simulations are conducted for bubble trains and certain differences in terms of changing streamline
patterns are expected. We examined velocity patterns to identify the moment of the streamlines
pattern change. We chose two
representative capillary numbers as $Ca=0.47$ and $Ca=0.63$. 
One can
see in Fig. \ref{fig:streamlines:pattern} a clear transition between associated patterns. Among all
different simulation runs with different grids and initial conditions, the
transition occurs at $Ca=0.6-0.7$, see Fig. \ref{fig:streamlines:pattern}. The
transition is identified between regimes where there exists and does not exist a vortex in the
slug. Either all streamwise slug velocities have the same sign (no vortex), or have different signs
(vortex). It is indicated by \citet{heil-bretherton,kreutzer-reactors} that the bubble train flow
depends on the slugs and bubbles lengths. Thus, we expect the difference of bubble train simulations
and simulations of a semi-infinite air finger propagation \cite{heil-threedim}. The transition also
depends on the identification of the bubble velocity
which is influenced by spurious currents and interpolation errors.
\begin{figure}[ht]
\includegraphics[angle=90,width=\textwidth]{Figures/stream_ca47.eps}\\
\includegraphics[angle=90,width=\textwidth]{Figures/stream_ca63.eps}\\
\caption{Velocity vector maps for $Ca=0.47$ and $Ca=0.63$. One can see that there are no vortexes
created
in front of the bubble for $Ca=0.63$. For presented simulations the transition happens between
$Ca=0.47$ and $Ca=0.63$, which
is a different value from $Ca=0.69$ \cite{heil-threedim}. However, we performed a number of
different types of simulations (different force, width initialization). The critical transition
capillary number $\widehat{Ca}$ is not well fixed and varies from $\widehat{Ca}=0.6-0.7$.  
\label{fig:streamlines:pattern}}
\end{figure}

\subsection{Variation over bubble length}
The work of \citet{wang-non-circular} shows the variation of the bubble
radii along the
bubble. The bubble shape in terms of axial and diagonal radii is reconstructed for a number of
capillary numbers, Fig. \ref{fig:bubble:variation:capillaries}. One can see that for smaller
capillary numbers the diagonal radius exhibits a small jump near the rear bubble cap. That agrees
with results of \citet{wang-non-circular}. However, our simulations show smoother behavior of
the jump.  
\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{Figures/bubble_rad_axis.eps}\\
\includegraphics[width=0.8\textwidth]{Figures/bubble_rad_diag.eps}\\
\caption{Radii variations along the bubble for different capillary numbers in the plane $x=0$.
One can see that the diagonal radius shows the jump near the bubble rear cap. This agrees with the
VOF simulations of \citet{wang-non-circular}.Coordinate $z$ increases in the streamwise direction.
\label{fig:bubble:variation:capillaries}}
\end{figure}

Another interesting phenomenon regarding the bubble shape was indicated by \citet{heil-threedim}. In
their simulations the non-axisymmetric shape was observed even for larger capillary numbers $Ca>4$
near
the bubble front cap which after certain distance relaxes to have radii equal to each other. The
bubble shape was calculated for the large capillary number $Ca=6.43$, see Fig.
\ref{fig:bubble:ca:large}. However, even the diagonal radius shows the jump near the front tip of
the bubble and it resembles the shape given by \citet{heil-threedim}, but the difference is around
$1$ percent and can be explained by error in the linear interpolation used for the
interface
calculation. Therefore, we do not observe the symmetry breakage for $Ca>4$. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{Figures/bubble_ca_large.eps}
\caption{The bubble shape for the capillary number $Ca=6.43$ is presented. $z$ increases in the
streamwise direction. One can see that near the front tip (large $z$) the diagonal radius exhibits
a jump which is relaxed after to the axis radius. Even this resembles
the shape of the curves obtained by \citet{heil-threedim}, but the
difference between radii is around $1$ percent and can be attributed to
the linear interpolation used in the calculation of the interface.\label{fig:bubble:ca:large}}
\end{figure}

@@

The present simulations show that the corresponding critical capillary number is close to
$\widehat{Ca}=0.1$ that agrees with the work of \citet{wang-non-circular} and is different from the
results of \citet{heil-threedim}, $\widehat{Ca}=0.04$. However, in the range of the moderate
capillary numbers our results are closer to the results of \citet{heil-threedim} and exhibit
less noisy trend in radii values and a more accurate transition between non-axisymmetric ans
axisymmetric cases than results of \citet{wang-non-circular}. 
%For moderate capillary numbers $Ca\geq 0.2$, the radii are smaller than the literature results.
%This
%may be caused by the bubble train simulation or the influence of the binary-liquid model. 
\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{Figures/capillaries_comparison_heil.eps}\\
\includegraphics[width=0.8\textwidth]{Figures/capillaries_comparison_wang.eps}\\
\caption{The comparison for the axial and the diagonal radii
versus capillary numbers between present simulations and the results of \citet{heil-threedim}
(top), of \citet{wang-non-circular} (bottom). One can see that
the code mimics
behavior of the earlier published results.  \label{fig:capillary:comparison}} 
\end{figure}

\subsection{Relative velocity}
One of the differencies between the flow in tubes and square capillaries is that the flow has
different streamline patterns. In square microchannels, the main mass flow occurs through corners
\cite{heil-threedim,wang-non-circular}. This affects the mass flow significantly. One of the
characteristics of the mass flow is the normalized dependence of the relative velocity on the
capillary
number. The bubble always moves faster than the surrounding liquid. The following quantity is
defined as
relative velocity \cite{cerro-bubble-train}:
\begin{equation}
W=\frac{U_{\mathrm{bubble}}-U_{ls}}{U_{\mathrm{bubble}}},
\end{equation}
where $U_{\mathrm{bubble}}$ is the bubble interface velocity, and $U_{ls}$ is the liquid superficial
velocity. As far as the steady state is achieved one can take the liquid superficial velocity
exactly at the cross section in the middle between bubbles. The liquid superficial velocity is
defined as:
\begin{equation}
U_{ls}=\frac{\int_{A}{U_{liq}dA}}{A}.
\end{equation}
The relative velocity was calculated and compared with the results of \citet{wang-non-circular},
see Fig. \ref{fig:relative:velocity}. One can see an agreement between the simulations indicating
that
one of the mass transport characteristics can be captured accurately. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{Figures/relative_velocity.eps}
\caption{Relative velocity comparison between current simulations and simulations of
\citet{wang-non-circular}. One can see a qualitative agreement. \label{fig:relative:velocity}}
\end{figure}

@@

\subsection{Rectangular channel simulations}
While the main focus of this paper is on the square microchannel simulations, we performed a number
of simulations for rectangular channels with varying aspect ratio
$\alpha=\frac{W}{H_{\mathrm{eff}}}=\frac{a_x}{a_y}$, where $a_x$ and $a_y$ are sides of the rectangular in $x$ and $y$ directions.\citet{heil-threedim} performed a number of simulations for the
propagation of semi-infinite air finger in the microchannel. They indicated that the radii in $x$
and $y$ directions  are increasing with the aspect ratio $\alpha$
increases. The same trend can be seen for the current simulation results, see Fig.
\ref{fig:aspect:ratio}. One can see that diagonal and axial radii are not the same as they were in square channels for the capillary number range investigated here. The same is
indicated by \citet{heil-threedim} who mentioned that for radii to coincide one needs to have really long bubbles, at least $100$ shorter semi-axis bubble length for the aspect ratio $\alpha=1.5$. This is not possible to achieve with the
current bubble train simulations.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/rectangle.eps}
\caption{Radii in $x$ ($R_x$) and $y$ ($R_y$) directions for rectangular microchannels with
different aspect ratios. Radii are rescaled on the length of channel in $y$ direction. One can see that with
the increase of $\alpha$ the radii increase as well. However, $R_x\neq R_y$. It was indicated by \citet{heil-threedim} one needs semi-infinite bubble to achieve $R_x=R_y$. \label{fig:aspect:ratio}}
\end{figure}

>>>>

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

@@

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which would require a
simultaneous imposement of pressure boundary conditions and the body force. 
%The pressure boundary
%conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
%horizontal direction only and the upward/downward flows with pressure difference imposed are not
%simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

@@

To further examine the steady-state, the velocity in the streamwise direction is plotted. The values
of the velocity are taken on the contour where $\phi=0$. This corresponds
 to the bubble interface. That allows to check whether the bubble is propagating as a whole rigid
body or it's shrinking or elongating. This is one of the characteristics of the steady-state.  One
can see in Fig. \ref{fig:velocity:contour} the contour plot and the
corressponing streamwise velocity
component. The crossection is a plane $x=0$, where $x$ points
towards the reader, $z$ is the streamwise direction. As fluxes inside the bubble have
clockwise and counterclockwise directions, see Section \ref{sec:velocity}, one needs to thoroughly
examine only the points corresponding to the front and the rear bubble caps. If they are equal, then
it is exactly the velocity with which the bubble propagates in the microchannel as a rigid body.
For a physical analogy, one can imagine the rotating ball propagating in the streamwise direction.
The front and rear ball points have the same streamwise velocity. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{velocity_interface_contour.eps}\\
\includegraphics[width=\textwidth]{velocity_interface_values.eps}\\
\caption{The bubble shape and the corresponding streamwise velocities taken on the bubble
interface after $240000$ iterations. Directions $y$ and $z$ are scaled to the half of the channel
$H_{\mathrm{eff}}/2$. One can see that the bubble front and rear caps are propagating with nearly
the
same velocity, that is the actual bubble velocity. For more details on velocity patterns
see Section \ref{sec:velocity}. \label{fig:velocity:contour}}
\end{figure}

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

\subsection{Velocity pattern}
\label{sec:velocity}
The liquid velocity pattern is known to change its behavior depending on the capillary number. For
relatively low capillary numbers $Ca<0.6$, in a reference frame moving with the bubble a vortex 
is observed \cite{heil-threedim}. For larger capillary numbers it is indicated that there is no
vortex in front of the bubble. The transition capillary number specified by
\citet{heil-threedim} for a square channel as $Ca=0.691$. However, the
distribution of vorticity strongly depends on the Reynolds number \cite{heil-bretherton}, as well as
on
the
pressure distribution. In the case of the bubble train it is indicated by \citet{kreutzer-taylor}
that the pressure is significantly influenced by bubble frequency and slug distance. The present
simulations are conducted for bubble trains and certain differences in terms of changing streamline
patterns are expected. We examined velocity patterns to identify the moment of the streamlines
pattern change. We chose two
representative capillary numbers as $Ca=0.47$ and $Ca=0.63$. 
One can
see in Fig. \ref{fig:streamlines:pattern} a clear transition between associated patterns. Among all
different simulation runs with different grids and initial conditions, the
transition occurs at $Ca=0.6-0.7$, see Fig. \ref{fig:streamlines:pattern}. The
transition is identified between regimes where there exists and does not exist a vortex in the
slug. Either all streamwise slug velocities have the same sign (no vortex), or have different signs
(vortex). It is indicated by \citet{heil-bretherton,kreutzer-reactors} that the bubble train flow
depends on the slugs and bubbles lengths. Thus, we expect the difference of bubble train simulations
and simulations of a semi-infinite air finger propagation \cite{heil-threedim}. The transition also
depends on the identification of the bubble velocity
which is influenced by spurious currents and interpolation errors.
\begin{figure}[ht]
\includegraphics[angle=90,width=\textwidth]{stream_ca47.eps}\\
\includegraphics[angle=90,width=\textwidth]{stream_ca63.eps}\\
\caption{Velocity vector maps for $Ca=0.47$ and $Ca=0.63$. One can see that there are no vortexes
created
in front of the bubble for $Ca=0.63$. For presented simulations the transition happens between
$Ca=0.47$ and $Ca=0.63$, which
is a different value from $Ca=0.69$ \cite{heil-threedim}. However, we performed a number of
different types of simulations (different force, width initialization). The critical transition
capillary number $\widehat{Ca}$ is not well fixed and varies from $\widehat{Ca}=0.6-0.7$.  
\label{fig:streamlines:pattern}}
\end{figure}

\subsection{Variation over bubble length}
The work of \citet{wang-non-circular} shows the variation of the bubble
radii along the
bubble. The bubble shape in terms of axial and diagonal radii is reconstructed for a number of
capillary numbers, Fig. \ref{fig:bubble:variation:capillaries}. One can see that for smaller
capillary numbers the diagonal radius exhibits a small jump near the rear bubble cap. That agrees
with results of \citet{wang-non-circular}. However, our simulations show smoother behavior of
the jump.  
\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{bubble_rad_axis.eps}\\
\includegraphics[width=0.8\textwidth]{bubble_rad_diag.eps}\\
\caption{Radii variations along the bubble for different capillary numbers in the plane $x=0$.
One can see that the diagonal radius shows the jump near the bubble rear cap. This agrees with the
VOF simulations of \citet{wang-non-circular}.Coordinate $z$ increases in the streamwise direction.
\label{fig:bubble:variation:capillaries}}
\end{figure}

Another interesting phenomenon regarding the bubble shape was indicated by \citet{heil-threedim}. In
their simulations the non-axisymmetric shape was observed even for larger capillary numbers $Ca>4$
near
the bubble front cap which after certain distance relaxes to have radii equal to each other. The
bubble shape was calculated for the large capillary number $Ca=6.43$, see Fig.
\ref{fig:bubble:ca:large}. However, even the diagonal radius shows the jump near the front tip of
the bubble and it resembles the shape given by \citet{heil-threedim}, but the difference is around
$1$ percent and can be explained by error in the linear interpolation used for the
interface
calculation. Therefore, we do not observe the symmetry breakage for $Ca>4$. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{bubble_ca_large.eps}
\caption{The bubble shape for the capillary number $Ca=6.43$ is presented. $z$ increases in the
streamwise direction. One can see that near the front tip (large $z$) the diagonal radius exhibits
a jump which is relaxed after to the axis radius. Even this resembles
the shape of the curves obtained by \citet{heil-threedim}, but the
difference between radii is around $1$ percent and can be attributed to
the linear interpolation used in the calculation of the interface.\label{fig:bubble:ca:large}}
\end{figure}

@@

The present simulations show that the corresponding critical capillary number is close to
$\widehat{Ca}=0.1$ that agrees with the work of \citet{wang-non-circular} and is different from the
results of \citet{heil-threedim}, $\widehat{Ca}=0.04$. However, in the range of the moderate
capillary numbers our results are closer to the results of \citet{heil-threedim} and exhibit
less noisy trend in radii values and a more accurate transition between non-axisymmetric ans
axisymmetric cases than results of \citet{wang-non-circular}. 
%For moderate capillary numbers $Ca\geq 0.2$, the radii are smaller than the literature results.
%This
%may be caused by the bubble train simulation or the influence of the binary-liquid model. 
\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{capillaries_comparison_heil.eps}\\
\includegraphics[width=0.8\textwidth]{capillaries_comparison_wang.eps}\\
\caption{The comparison for the axial and the diagonal radii
versus capillary numbers between present simulations and the results of \citet{heil-threedim}
(top), of \citet{wang-non-circular} (bottom). One can see that
the code mimics
behavior of the earlier published results.  \label{fig:capillary:comparison}} 
\end{figure}

\subsection{Relative velocity}
One of the differencies between the flow in tubes and square capillaries is that the flow has
different streamline patterns. In square microchannels, the main mass flow occurs through corners
\cite{heil-threedim,wang-non-circular}. This affects the mass flow significantly. One of the
characteristics of the mass flow is the normalized dependence of the relative velocity on the
capillary
number. The bubble always moves faster than the surrounding liquid. The following quantity is
defined as
relative velocity \cite{cerro-bubble-train}:
\begin{equation}
W=\frac{U_{\mathrm{bubble}}-U_{ls}}{U_{\mathrm{bubble}}},
\end{equation}
where $U_{\mathrm{bubble}}$ is the bubble interface velocity, and $U_{ls}$ is the liquid superficial
velocity. As far as the steady state is achieved one can take the liquid superficial velocity
exactly at the cross section in the middle between bubbles. The liquid superficial velocity is
defined as:
\begin{equation}
U_{ls}=\frac{\int_{A}{U_{liq}dA}}{A}.
\end{equation}
The relative velocity was calculated and compared with the results of \citet{wang-non-circular},
see Fig. \ref{fig:relative:velocity}. One can see an agreement between the simulations indicating
that
one of the mass transport characteristics can be captured accurately. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{relative_velocity.eps}
\caption{Relative velocity comparison between current simulations and simulations of
\citet{wang-non-circular}. One can see a qualitative agreement. \label{fig:relative:velocity}}
\end{figure}

@@

\subsection{Rectangular channel simulations}
While the main focus of this paper is on the square microchannel simulations, we performed a number
of simulations for rectangular channels with varying aspect ratio
$\alpha=\frac{W}{H_{\mathrm{eff}}}=\frac{a_x}{a_y}$, where $a_x$ and $a_y$ are sides of the rectangular in $x$ and $y$ directions.\citet{heil-threedim} performed a number of simulations for the
propagation of semi-infinite air finger in the microchannel. They indicated that the radii in $x$
and $y$ directions  are increasing with the aspect ratio $\alpha$
increases. The same trend can be seen for the current simulation results, see Fig.
\ref{fig:aspect:ratio}. One can see that diagonal and axial radii are not the same as they were in square channels for the capillary number range investigated here. The same is
indicated by \citet{heil-threedim} who mentioned that for radii to coincide one needs to have really long bubbles, at least $100$ shorter semi-axis bubble length for the aspect ratio $\alpha=1.5$. This is not possible to achieve with the
current bubble train simulations.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{rectangle.eps}
\caption{Radii in $x$ ($R_x$) and $y$ ($R_y$) directions for rectangular microchannels with
different aspect ratios. Radii are rescaled on the length of channel in $y$ direction. One can see that with
the increase of $\alpha$ the radii increase as well. However, $R_x\neq R_y$. It was indicated by \citet{heil-threedim} one needs semi-infinite bubble to achieve $R_x=R_y$. \label{fig:aspect:ratio}}
\end{figure}

They as well indicated that the simulations results with different aspect ratio can be
put on one curve by calculating non-dimensional radius $s_{\infty}$ that scales different aspect
ratio microchannels. It is based on the calculation of  $r_{\infty}$ (the radius that the air
finger will have at an infinite distance from the bubble tip):
\begin{equation}
\begin{aligned}
&\frac{r_{\infty}}{\frac{H_{\mathrm{eff}}}{2}}=\frac{\sqrt{\frac{Q}{\pi}}}{\frac{H_{\mathrm{eff}}}{2
} }\\
&s_{\infty}=\frac{r_{\infty}}{\frac{H_{\mathrm{eff}}}{2}}\alpha^{-1/2},
\end{aligned}
\end{equation}
where $Q$ is the area occupied by air in the crossection. \citet{heil-threedim} indicated a simple
physical interpretation of this phenomena:  providing that air is injected with the same
velocity to the channels with different aspect ratio but the same area, the resulting occupied area by the bubble
should not depend on the aspect ratio. That is quite interesting phenomena indicating that if bubbles are infinite long then they would have the same radius and will be circular. The current results, see Fig. \ref{fig:one:curve} show that
this hypothesis can be extended for bubble train flows with different radii $R_h \neq R_v$. 
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{onecurve.eps}\\
\caption{The non-dimensional radius $s_{\infty}$ for different aspect ratio
rectangular microchannels. One can see that all the curves coincide
independently of the aspect ratio.\label{fig:one:curve}}
\end{figure}
